---
title: Scraping online table with info on R datasets
author: Kasia Kulma
date: '2017-03-30'
slug: scraping-online-table-with-info-on-r-datasets
categories:
  - r
tags:
  - web scraping
description: ''
topics: []
---

It's a very quick post on how to get a list of datasets available from within R with their basic description (what package they can be found in, number of observations and variables). It always takes me some time to find the right dataset to showcase whatever process or method I'm working with, so this was really to make my life easier. So! I'm going to scrape the table with a list of R datasets from [here](https://vincentarelbundock.github.io/Rdatasets/datasets.html) using `rvest` and `xml2` packages:

```{r, warning=FALSE, message=FALSE}
# loading packages
library(rvest)
library(xml2)
library(dplyr)
library(knitr)

# URL to scrape
url <- "https://vincentarelbundock.github.io/Rdatasets/datasets.html"

# scrape the table with relevant info
r_datasets <- read_html(url) %>% # read url
    html_nodes("table") %>% # extract all the tables
   .[[2]] %>% # it's the second table we want
    html_table() # convert it to a usable format (data.frame)
```

As a result, we get a tidy data frame...

```{r}
str(r_datasets)
```


```{r}
r_datasets %>% 
  select(-c(CSV, Doc)) %>% 
  head() 
```

.. that we can filter freely, according to our needs:

```{r, warning=FALSE, message=FALSE}
r_datasets %>% filter(Rows >= 1000 & Cols >= 50) %>% 
  kable()
```

```{r}
r_datasets %>% filter(grepl("cat", Item)) %>% kable()
```

This totally maked my life easier, so hope it will help you, too!
