---
title: My R take on Advent of Code - Day 2
author: Kasia Kulma
date: '2018-12-19'
categories:
  - r
tags:
  - tidyverse
slug: my-r-take-on-advent-of-code-day-2
---

This is my second blog post from the series of `My R take on Advent of Code`. If you'd like to know more about Advent of Code, check out the [first post from the series](https://r-tastic.co.uk/post/advent-of-code-in-r-day-1/) or simply go to [their website](https://adventofcode.com/2018).  Below you'll find the challnge from Day 2 and the solution that worked for me. As always, feel free to leave comments if you have different ideas on how this could have been solved!


### Day 2 Puzzle


>> (...) you scan the likely candidate boxes again, counting the number that have an ID containing exactly two of any letter and then separately counting those with exactly three of any letter. You can multiply those two counts together to get a rudimentary checksum and compare it to what your device predicts.
>> For example, if you see the following box IDs:

>> `abcdef` contains no letters that appear exactly two or three times.    
>> `bababc` contains two `a` and three `b`, so it counts for both.    
>> `abbcde` contains two `b`, but no letter appears exactly three times.    
>> `abcccd` contains three `c`, but no letter appears exactly two times.    
>> `aabcdd` contains two `a` and two `d`, but it only counts once.    
>> `abcdee` contains two `e`.    
>> `ababab` contains three `a` and three `b`, but it only counts once.   

>> Of these box IDs, four of them contain a letter which appears exactly twice, and three of them contain a letter which appears exactly three times. Multiplying these together produces a checksum of 4 * 3 = 12.
>> What is the checksum for your list of box IDs?

So what is it all about? As complicated as it may sound, essentially we need to:

- understand which string contains letters that appear exactly 2 times
- understand which string contains letters that appear exactly 3 times
- count the number of each type of string
- multiply them together 

Doesn't sound so bad anymore, ey? This is how we can go about it:

First load your key packages...

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(stringr)
library(tibble)
library(purrr)
```

```{r, echo = FALSE}
input <- 'xrecqmdonskvzupalfkwhjctdb
xrlgqmavnskvzupalfiwhjctdb
xregqmyonskvzupalfiwhjpmdj
areyqmyonskvzupalfiwhjcidb
xregqpyonskvzuaalfiwhjctdy
xwegumyonskvzuphlfiwhjctdb
xregumymnskvzupalfiwhjctib
xregqmyonjkvzupalfvwijctdb
xrmgqmyonsdvzupalfiwhjcthb
xrpgqmyonskvzupalfiwhcitdb
xregvmysnsjvzupalfiwhjctdb
xregqsyonskvzupgqfiwhjctdb
qreuqmyonskvzupalfiwyjctdb
xrecqmyenskvzupalyiwhjctdb
xmegqmyonskvzhpalfcwhjctdb
xiegqmyonskvzupalfkwhjjtdb
xreaqmyofskfzupalfiwhjctdb
xregqmypnskvzupalmiwhjltdb
xretqmmonskvzupalfiwhwcfdb
xrexqmkonskvzupalfiwjjctdb
xrigqmyonskvgupplfiwhjctdb
xregqmyotskvzupalfywhjctdm
xcegmmyonsvvzupalfiwhjctdb
xrezqmypnskvznpalfiwhjctdb
xragqmyonskvzupblfiwajctdb
xregqmyonskvzwpapfiwhjctqb
xoegqmyoyskvzupaufiwhjctdb
xrcgqmyjnskvzupalfcwhjctdb
xregqmyonskvzudalfipajctdb
xsegqmyonsklzupalwiwhjctdb
xregqmyocskvduaalfiwhjctdb
xhegqmyfnskvzupalflwhjctdb
xregqmymnykvzupalfiwhjctdm
xregqmybnskvzupacfiwdjctdb
xaegqmlonskvzfpalfiwhjctdb
xoegtmyonskvzupalfiwhwctdb
xregqmyohskvzupaqfiwhjccdb
xoegqmyonstvzupalfiwhjctbb
mregnmyonskszupalfiwhjctdb
xreoqmycnskvzupalfiphjctdb
xregqmyocskvdupacfiwhjctdb
xregqmyonskvzupajqiahjctdb
xregqmyonslvwupalfiwhjcfdb
xregqmyonskvzapalfiwhqcthb
xrerqmyonskwzupalfiwhjctdt
xrefqmfonskvzupalfiwcjctdb
xregqmyonskvzupadfiwhjxedb
iregqhyonskvzupaliiwhjctdb
iregqmyotskvzucalfiwhjctdb
xrbgqmaonrkvzupalfiwhjctdb
xregqmyonskvzupalfixhdctdf
xrehqmyonskvzupalfiwijctdd
xvegqmyonskvzupaleuwhjctdb
xregqmyiyskvzupalfiwqjctdb
hregqmyonskvzupaxfiwhjptdb
xregamyznskbzupalfiwhjctdb
xreyqmyonskvgupalziwhjctdb
xregqmysnskvzupalfiwhgctdu
xojgqmyonskvzupalfiwbjctdb
xrkgqmyonskvlupalfiwhjcwdb
xregqmyonwkvxupalfiwajctdb
xregqmyonsuvzupalfjwhjcxdb
xregqmyonskgzucalfiwhjstdb
xaegqmyonfkvzupalfiwhjcttb
xlegqmyonskvzupazfiwhjctqb
xrejqmyonskvzqpaldiwhjctdb
xreguryonskvzupalfiwhjctdz
xregqsyoeskvzupalfiwhjctdt
xregqmyonskvzubalfirhjctdp
xrepqmymnskvzupadfiwhjctdb
xregqayonskvzuoalfichjctdb
xreqqmyonskvzunalfiwojctdb
xregqmyonsivzufalciwhjctdb
xregqqeonskvzupanfiwhjctdb
xoegqmyunskvzppalfiwhjctdb
xregqmyonskvzupalfqwhnftdb
xregqmyonskvzuralkiwhjcudb
xrwgqmymnskvzupalfiwhjcgdb
xvrgqmyonskvzupalfiwhjcthb
xregemyonskkzupalfiwhjctbb
xregqmyonsevzupalfiwhjjtdl
xregqmyonckvcupajfiwhjctdb
xregqmysnskvzunalfnwhjctdb
xreowmyonskvkupalfiwhjctdb
xregqmyonskvjupalfiwhjytdr
xregqmyonskyzupaffiwhmctdb
xrsgqmyonszvzupmlfiwhjctdb
xzegqmyonskvnupalfiwfjctdb
qregqmyonskvzupalfiwhrctjb
xpegqmyonsivzupqlfiwhjctdb
xregqmyoyskrzupalfiwhjctdx
xregqmyonsqvzupalfiwhjdndb
xregjmyonskvzppalfiwhjcgdb
xregqmyziskvzupalfiwhjctib
xregqmyonmkvbupalfiwhjckdb
xtegamyonskvzupalniwhjctdb
xregqpyonskvzhpwlfiwhjctdb
xvegqmfonskvzupalfiwhjcadb
xregqmyonskvzupaysiwhjctxb
xrejqmyonudvzupalfiwhjctdb
llegqmyonskvzbpalfiwhjctdb
tcegqmbonskvzupalfiwhjctdb
lregqmyohskvzupalfiwhjcttb
xrngqmcfnskvzupalfiwhjctdb
xregqmyonspvzuuplfiwhjctdb
xrxgqmyonslvzupalfiwhjctdo
xregqmyonskvzulalfuwhjdtdb
xregqmnonskvzupalfvwhjckdb
xregqbyfnskvzupaltiwhjctdb
xregqmyodsovzwpalfiwhjctdb
xregomyonskvhrpalfiwhjctdb
xregqmfdnskvzupalliwhjctdb
xregqmyonskvzupaabithjctdb
xrngamyonskvzupalfiwhjcttb
xrhgqmyonskvzupaldifhjctdb
xrygzmyonskvzupatfiwhjctdb
xregqmyonskvzupiqtiwhjctdb
xregqmyonfkvzupalfiwxjcsdb
xregqsyunskvzupalfiwhjctde
xrzgqmyolskvzupasfiwhjctdb
xgegqmyoyskvzupalfiwfjctdb
xrvgqlyohskvzupalfiwhjctdb
xregcmyonskvzuprlyiwhjctdb
xregqmyonskvwjpalfiwsjctdb
xrfgqmyonskvzupalfidhactdb
xcegqmyonwkvzdpalfiwhjctdb
nregqmyrnskvzupalciwhjctdb
xcegqmyonskvzvpalfiwhjctdj
xregqmyonskvzupqssiwhjctdb
xregcmyonskvzupalfinhjutdb
xregqmyonskvzupzlfiwcjctnb
xnegqmyozskvzbpalfiwhjctdb
xregvmponskvzupalfiwhsctdb
xregqmyonskvpupalqichjctdb
xreqqmyonskvzupauuiwhjctdb
xregqryonskvzupatfiwhjctyb
hregqmyonokvzupalfiwhmctdb
xreuqmionckvzupalfiwhjctdb
xregqmyoiskvzupanfiwhjntdb
xrdgqmronskvzupaluiwhjctdb
xadgqmyunskvzupalfiwhjctdb
eregqmzonskvzupakfiwhjctdb
xiegqmyonskvnupblfiwhjctdb
yregqmzonskvzupalfiwhjotdb
xregqmyonskvjupalfiwhjhtvb
wregqmyonskvzzprlfiwhjctdb
xregqmyovskvzupalgiuhjctdb
xregqmyonskjzupelfuwhjctdb
xregqmysuskvpupalfiwhjctdb
xrebqkyonskvzupalfiwpjctdb
xregcmyonskvzipalfiwhjcttb
xregqmyonskdyupalfiwgjctdb
xregcmyonskvzupalfiwijctnb
xregqmyonsovdupalfrwhjctdb
xregqmaonskvzupalnkwhjctdb
xregqmysnfkvzupalfiwhictdb
xregqmyonswvzupalfiyhjctdf
xreoqmyrnskvzupalfihhjctdb
tregqmydnskvzupalfizhjctdb
xregxmyonykvzupalfnwhjctdb
xzegqnyonskuzupalfiwhjctdb
xregqmfonszvvupalfiwhjctdb
xrerqmyjnskvzupalfiwhpctdb
xregqmyanskvzupalffphjctdb
rregqmyogskvzupalfiehjctdb
xrpgqmyonspvzupalfiwgjctdb
xuegqmppnskvzupalfiwhjctdb
xregqmyonskvzqpalsiwhjhtdb
xregqzyonskvzkpalfiwujctdb
xrdgqmyonskvzupglfiwhjctdu
xregqmyonskqzupahciwhjctdb
treqqmyonskvzupalfiwhjcqdb
vlegqmyonskvzupalfiwhjwtdb
xregjmyonskviupglfiwhjctdb
xreggmyanskvzupalfiwhjcydb
xregqmybnskvzuprlfiwhjmtdb
xrsgqmyonskizupagfiwhjctdb
xregqmyenskvzupalfvwhjctib
lrygqmyonsrvzupalfiwhjctdb
xregqmjonskvqupalfiwhjctdu
xregqmyonsknzmpzlfiwhjctdb
xregqmyonhkvzupllfiwhjctdz
xregqmronskvdumalfiwhjctdb
xrpgqmyonskvzupalfhwhjhtdb
xfegqmeonskvzupasfiwhjctdb
xregqqyonskvzrpalfiwijctdb
xretqmmonskvzupalfiwhjcfdb
xregqmyonskvznpalniwhjztdb
xregqmyqnskvzuoalfiwhhctdb
xregqmyonsbvzupalviwhjxtdb
xregqmyonskvzupazmiwhhctdb
xregqmyosskvzupalflwhjctdw
xtegqmyonskvzupamciwhjctdb
xregamyonskvzbpalfiwhqctdb
xregqmgonskvzupalfiwhictxb
xregqmyonskvjupvlfnwhjctdb
xrthqmyonskvzupalfiwhjctub
xrexqmyoyskvzupalfiwhjcadb
xvegqmyonskvxupalfiwhjztdb
xregqmyonskgzupalhiwhjptdb
xregqmysnskvzufalpiwhjctdb
xregqmyonskvbipalfighjctdb
xregqmyonskvzupylfiwhjwvdb
gregqmyonskvzupalfikhjctdt
ujegqmyonskvzupalfiwhjctlb
nreqqmyonskjzupalfiwhjctdb
xregqmyonskvzupanfbwhjchdb
xregqyyoeskwzupalfiwhjctdb
xregqmyokskvzgpalfiwhnctdb
lregqmyonskvzupalfawsjctdb
xtegqmyonskvzmpalfiwhjctmb
xtegqvyonskvzupalfiwhjdtdb
xpegqpyonekvzupalfiwhjctdb
qregqmyonskvzupalfiwmjctdn
xregqnyosskvzupalfibhjctdb
xregqmyonsknzupalflwhjctfb
xregqmxoyskvzuealfiwhjctdb
xregdmyoeskvzupalfiwhfctdb
xremmmyonskvzupalfiwhxctdb
xregqmconskvzupylfuwhjctdb
xregqmyonskvzupawiiwhictdb
xlegsmyonskvzupalfiwhbctdb
xregqmyonsavzopalyiwhjctdb
xregqmyonskczupalfibhvctdb
xregqmyonskvzvpalfiunjctdb
xregqmyonskvdupalfiwhjczdp
xregqmyonskvzupklfswhhctdb
xrelqmyonskvzupalyiwhjctdi
xrcgqmyonskvzupalfieqjctdb
xregqmnonskvzupacfewhjctdb
xrwgqmyonskvzuealfiwhcctdb
xregqiyonsevzmpalfiwhjctdb
xregqmyonjyvzupalfiwhjckdb
xregqmyonyklzupadfiwhjctdb
xregqmyanskvzupolfiwhjctpb
xdbgqmyonskvzupslfiwhjctdb
xregqmhonykvzupalfawhjctdb
xregqmqonsivzupalfifhjctdb
xregqgyonsrvzupalfiwhjctib
xregqmyofskvzupalfiwlfctdb
xregqmyovskvzupllftwhjctdb
xregqmyonskvzupaciiwhuctdb
xregqmyonsdvzuhalfiwhjhtdb
xreiqmyonskvzupalfiwhncldb
xregqmyongkvzugalfiwhjctxb
xregqsyonskvzrpmlfiwhjctdb
xrogqmyonskvzxpalfiwhbctdb
xregqmkonskvzuqalfiwhjptdb
xregqmyonskvvxpalfiwhactdb
xregqmyonskvzupsliiwhwctdb'
```

... and have a look at what the raw input looks like.

```{r}
# check raw input
glimpse(input)
```

Right, Advent of Code will never give you nice and clean data to work with, that's for sure. But it doesn't look like things are too bad this time - let's just split it by the new line and keep it as a vector for now. Does it look reaosnably good?

```{r}
# clean it
clean_input =  strsplit(input, '\n') %>% unlist()   # splt by NewLine
glimpse(clean_input)
```

Much better! Now, let's put it all in a data frame for now, we'll need it very soon.  

```{r}
# put it in the data.frame
df2 <- tibble(input = str_trim(clean_input))
head(df2)
```

Now, the way I approached this was to split each word into letters and then count how many times they occured. Then, for identifying words with 2 occurences, I filtered only those that occur twice and if the final table has any rows, then this counts as yes. Take the first example:

```{r}
strsplit(input, '\n') %>% unlist() %>% .[[1]] # get the first example
```

Let's split it by the letter, put it in a tibble and count each letter occurances:

```{r}
strsplit(input, '\n') %>% unlist() %>% .[[1]] %>% # get the first example 
  strsplit('') %>% # split letters
  unlist() %>% # get a vector
  as_tibble() %>% # trasform vector to tibble
  rename_(letters = names(.)[1]) %>% # name the column: letters 
  count(letters)
```

Now, do we have any double occurances there? 

```{r, warning=FALSE, message=FALSE}
# test: counting double letter occurances 
strsplit(input, '\n') %>% unlist() %>% .[[1]] %>% # get the first example 
  strsplit('') %>% # split letters
  unlist() %>% # get a vector
  as_tibble() %>% # trasform vector to tibble
  rename_(letters = names(.)[1]) %>% # name the column: letters 
  count(letters) %>% # count letter occurances
  filter(n == 2) %>% # get only those with double occurances
  nrow() # how many are there?
```

Definitely yes. Let's repeat the process for tripple occurances:


```{r, warning=FALSE, message=FALSE}
# test: counting triple letter occurances 
strsplit(input, '\n') %>% unlist() %>% .[[1]] %>% # get the first example
  strsplit('') %>% # split letters
  unlist() %>% 
  as_tibble() %>% # trasforming vector to tibble
  rename_(letters = names(.)[1]) %>% 
  count(letters) %>% 
  filter(n == 3) %>% 
  nrow()
```

Not much luck with those in this case. To make our life easier, let's wrap both calculations in functions...

```{r}
### wrap-up in functions
# count double occurances 
count2 <- function(x) {
  result2 <-  as.character(x) %>% 
    strsplit('') %>% # split by letters
    unlist() %>% 
    as_tibble() %>% # trasforming vector to tibble
    rename_(letters = names(.)[1]) %>% 
    count(letters) %>% # count letter occurances
    filter(n == 2) %>% 
    nrow()
  return(result2)
}


# count triple occurances 
count3 <- function(x) {
  result2 <-  as.character(x) %>% 
    strsplit('') %>% 
    unlist() %>% 
    as_tibble() %>% # trasforming vector to tibble
    rename_(letters = names(.)[1]) %>% 
    count(letters) %>% 
    filter(n == 3) %>% 
    nrow()
  return(result2)
}
```

...and apply them to the whole dataset:

```{r}
### apply functions to input
occurs2 <- map_int(df2$input, count2)
occurs3 <- map_int(df2$input, count3)
str(occurs2)
```

Now, all we need to do is check how many positive elements we have in each vector and multiple their lengths by each other:


```{r}
#solution
length(occurs2[occurs2 != 0]) * length(occurs3[occurs3 != 0])
```

Voila!