---
title: My R Take in Advent of Code - Day 5
author: Kasia Kulma
date: '2019-01-05'
slug: my-r-take-in-advent-of-code-day-5
categories:
  - r
tags:
  - tidyverse
description: ''
topics: []
draft: true
---

>> The polymer is formed by smaller units which, when triggered, react with each other such that two adjacent units of the same type and opposite polarity are destroyed. Units' types are represented by letters; units' polarity is represented by capitalization. For instance, r and R are units with the same type but opposite polarity, whereas r and s are entirely different types and do not react.

>> For example:
In **aA**, a and A react, leaving nothing behind.
In **abBA**, bB destroys itself, leaving aA. As above, this then destroys itself, leaving nothing.
In **abAB**, no two adjacent units are of the same type, and so nothing happens.
In **aabAAB**, even though aa and AA are of the same type, their polarities match, and so nothing happens.
Now, consider a larger example, **dabAcCaCBAcCcaDA**:

>> **dabAcCaCBAcCcaDA**  The first 'cC' is removed.    
**dabAaCBAcCcaDA**    This creates 'Aa', which is removed.    
**dabCBAcCcaDA**      Either 'cC' or 'Cc' are removed (the result is the same).    
**dabCBAcaDA**        No further actions can be taken.    

>> After all possible reactions, the resulting polymer contains **10 units**.
>> How many units remain after fully reacting the polymer you scanned?

```{r, warning=FALSE, message=FALSE}
library(tidyverse)

raw_input <- read_delim('day5-raw-input.txt', delim = '\t', col_names = FALSE) %>% 
  as.character()

glimpse(raw_input) 
```


```{r, warning=FALSE, message=FALSE}
# original length
nchar(raw_input) 
``` 


```{r, warning=FALSE, message=FALSE}
pattern <- c('Aa|aA|Bb|bB|Cc|cC|Dd|dD|Ee|eE|Ff|fF|Gg|gG|Hh|hH|Ii|iI|Jj|jJ|Kk|kK|Ll|lL|Mm|mM|Nn|nN|Oo|oO|Pp|pP|Qq|qQ|Rr|rR|Ss|sS|Tt|tT|Uu|uU|Vv|vV|Ww|wW|Xx|xX|Yy|yY|Zz|zZ') 


# test the example 

## dabAcCaCBAcCcaDA  The first 'cC' is removed.
#dabAaCBAcCcaDA    This creates 'Aa', which is removed.
#dabCBAcCcaDA      Either 'cC' or 'Cc' are removed (the result is the same).
#dabCBAcaDA 

  
str_remove_all('dabAcCaCBAcCcaDA', pattern) %>% 
  str_remove_all(pattern) %>% nchar()
```


```{r, warning=FALSE, message=FALSE}
# now, put it in the repeat loop
# initialize

pattern <- c('Aa|aA|Bb|bB|Cc|cC|Dd|dD|Ee|eE|Ff|fF|Gg|gG|Hh|hH|Ii|iI|Jj|jJ|Kk|kK|Ll|lL|Mm|mM|Nn|nN|Oo|oO|Pp|pP|Qq|qQ|Rr|rR|Ss|sS|Tt|tT|Uu|uU|Vv|vV|Ww|wW|Xx|xX|Yy|yY|Zz|zZ')


input <- raw_input
repeat {   
  output <- str_remove_all(input, pattern)
  if (nchar(input) == nchar(output)) {
    print(nchar(output));
    break
  } else input <- output;
}
```